---
title: "Building Tidy R Packages"
#institute: "R/Pharma workshop 2021"
author: "Leigh Alexander | Juliane Manitz"
date: "R/Pharma workshop | October 27, 2021"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      countIncrementalSlides: no
      highlightStyle: github
      titleSlideClass:
       - right
       - bottom
    self_contained: no
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
#options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(message = FALSE, warning = FALSE, cache = FALSE, 
                      error = TRUE, dev = "cairo_pdf", fig.retina = 2)

library(devtools)
library(usethis)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
#?style_mono_accent

style_mono_accent(
  base_color = "#1c5253", 
  inverse_text_color = "#cbf7ed",
)

style_extra_css(
  css = list(
    ".inverse h4" = list(color = "orange"),
    ".title-slide h1" = list(color = "orange")
  ))
```

```{css, include=FALSE}
/* custom.css */
.big-left {
  width: 60%;
  height: 92%;
  float: left;
}
.small-right {
  width: 37%;
  float: right;
  padding-left: 1%;
}

.small-left {
  width: 40%;
  height: 92%;
  float: left;
}
.big-right {
  width: 57%;
  float: right;
  padding-left: 1%;
}
.remark-code {
  font-size: 24px;
}
.small .remark-code { /*Change made here*/
  font-size: 80% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 50% !important;
}
```

# Workshop Description

**Why make an R package?** So many reasons! From sharing code and collaboration, to reproducibility and regulatory concerns, and finally for organizing you own analysis projects and tidy coding practices, R packages are here to help!

In this workshop, we will cover the very first steps of R package creation, the basic R package structure as well as all relevant components. 

-	Start to write R functions, 
-	document them with help pages, 
-	integrate examples and data, 
-	make sure that your code does what is expected by implementing unit testing, and 
-	finally built it and put it on CRAN or wherever you want to share it. 

We conclude the workshop with a selection of advanced topics related to R packages (e.g., package validation, objected-oriented programming/S3 methods or advanced package documentation with `pkgdown`). 

**Requirements:** This workshop requires some experience with R and is targeted at beginner to intermediate R users

---
# Introductions: Who are we? 

.pull-left[
add content about Leigh
]

.pull-right[
add content about Juliane
]
---
# Introductions: Who are you? 

1. Experience with R: ranges [beginner <intermediate < advanced]

2. Experience with R Package development: ranges [novice < some experience]

3. Which topics are you most interested (choose 3)? 
  - Basic Package Structure
  - Tools for R packaging
  - Unit testing
  - Advanced Documentation 
  - Package validation, 
  - Object orientated programming in R

4. What to you wanna use package development for? 
  - Personal use, 
  - internal company-use, 
  - make it publically available


---
class: inverse
# Workshop Outline: Part II

Basic R package components: 

- *DESCRIPTION* - What does the package? 

- *R/* - Start to write R functions, 

- *data/* - integrate examples and data, 

- *man/* - document them with help pages, 

- *tests/* - make sure that your code does what is expected by implementing unit testing, and 

- finally built it and put it on CRAN or wherever you want to share it. 

---
class: inverse
# Workshop Outline: Part II

Basic R package components: 

- *DESCRIPTION* - What does the package? 

- #### R/ - Start to write R functions, 

- *data/* - integrate examples and data, 

- *man/* - document them with help pages, 

- *tests/* - make sure that your code does what is expected by implementing unit testing, and 

- finally built it and put it on CRAN or wherever you want to share it. 

---
# Write R functions

- `R/` directory contains all R code
- A package with just an `R/` directory is still a very useful package.

--

### Good practice to structure:

- each function in a separate file (good for small packages)
- everything in one file (ok for small packages)
- group related functions in a file with meaningful names (best solution for larger projects)

---
# Workflow to Integrate R code

1. Edit your code.
2. Load your code with one of `devtools::load_all()`, which re-loads all saved files in `R/` into memory.
3. Experiment in the console.
4. Repeat

*Tipp*: Ctrl/Cmd + Shift + L (keyboard shortcut) saves all open files then calls `load_all()`.

---
class: inverse
# Workshop Outline: Part II

Basic R package components: 

- *DESCRIPTION* - What does the package? 

- *R/* - Start to write R functions, 

- #### data/ - integrate examples and data, 

- *man/* - document them with help pages, 

- *tests/* - make sure that your code does what is expected by implementing unit testing, and 

- finally built it and put it on CRAN or wherever you want to share it. 

---
# Package Example Data 

- The `data/` directory allows you to include data with your package.
- distribute 
- If `DESCRIPTION` contains `LazyData: true`, then the datasets won't occupy any memory until you use them

### Store data in

- `data/` to make data available to package users
- `R/sysdata.rda` to keep data internal for use by your functions.
- `inst/extdata` to make raw data available for loading and parsing examples. Access this data with `system.file()`

???

- Always use LazyData: true in your DESCRIPTION file.
- Store data in one of data/, R/Sysdata.rda, inst/extdata

---
# Workflow to Integrate Datasets

- Save data as `.Rdata` files in `data/`, which can be created by `save()`

- Make sure to adhere rules by using `devtools::use_data()`, which adds a data object to `data/` 

- data object gets added to `R/Sysdata.rda` if `internal = TRUE`

- `devtools::use_data_raw()` adds an R Script used to clean a data set to `data-raw/` and includes `data-raw/` on `.Rbuildignore`.

???

- Save data as .Rdata files (suggested)
- compress data to optimal value using option in `use_data()`

---
# Toy Example

The standard structure can be obtained automatically using `package.skeleton()`:

```{r, echo=TRUE}
# define some basic functions
add <- function(x, y){ x + y }
plusone <- function(x){ x + 1 }

# test your source code
add(10, 1)
plusone(4)

# prepare some example data
dat <- data.frame(id=1:10, x=rpois(10, 5), y=rpois(10, 5))

# create standard structure
fdlist <- c("add","plusone","dat")
utils::package.skeleton("myutils", fdlist, force=TRUE)
```

???

``` {r}
# test your source code
add(10, 1)
plusone(4)
dat 
```

---
class: inverse
# Workshop Outline: Part II

Basic R package components: 

- *DESCRIPTION* - What does the package? 

- *R/* - Start to write R functions, 

- *data/* - integrate examples and data, 

- #### *man/* - document them with help pages, 

- *tests/* - make sure that your code does what is expected by implementing unit testing, and 

- finally built it and put it on CRAN or wherever you want to share it. 

---
# R Function Documentation

- `man/` contains the documentation for your functions, the help pages in your package.
- R documentation format is very LaTeX-like output (LaTeX installation required)

.small[
```{r, eval=FALSE, echo=TRUE, size="small"}
\name{add}
\alias{add}
\title{Add together two numbers}
\usage{ add(x, y) }
\arguments{
  \item{x}{A number}
  \item{y}{A number}
}
\value{
  The sum of \code{x} and \code{y}
}
\description{ Add together two numbers }
\examples{
  add(1, 1)
  add(10, 1)
}
```
]

---
# Documentation with roxygen

- The `roxygen2` package lets you write documentation inline in your R-files
- `devtools` implements `roxygen2` to make documentation. 

### Workflow

1. Add roxygen comments in your `.R` files
2. Convert roxygen comments into documentation with `devtools::document()` which converts roxygen comments to `.Rd` files, places them in `man/.` and builds `NAMESPACE`.
3. Open help pages with `?` to preview documentation
4. Repeat

???

Ctrl/Cmd + Shift + D (Keyboard Shortcut)

---
# Documentation with roxygen

- roxygen comments start with `#'`
- Place comment lines directly above the code that defines the object documented.
- tags like `@param`, `@return`, `@author` define specific sections in `.Rd` file
- Untagged lines will be used to generate a title, description, and details section (in that order)

.small[
```{r, eval=FALSE, echo=TRUE}
#' Add together two numbers.
#'
#' @param x A number.
#' @param y A number.
#' @return The sum of \code{x} and \code{y}.
#' @examples
#' add(1, 1)
#' @export
add <- function(x, y) {
 x + y
}
```
]

???

---
# R help

!! integrate picture

---
# NAMESPACE

- The `NAMESPACE` file helps you make your package self-contained: it won’t interfere with other packages, and other packages won’t interfere with it.
- although the `NAMESPACE` file looks like R code, it is not processed as R code
- specifies which variables in the package should be **exported** to make them available to package users, and which variables
should be **imported** from other packages

``` {r eval=FALSE}
# import all functions from packages foo and bar
import(foo, bar) 
# import selected functions f and g from foo
importFrom(foo, f, g) 
# export functions f and g
export(f, g) 
```

---
# Create the NAMESPACE using roxygen 

- Export functions for users by placing `@export` in the roxygen comments
- Import objects from other packages with `package::object` (recommended) or 
- `@import`/`@importFrom`

--

### Workflow

1. Modify your code 
2. Document your package, e.g. using `devtools::document()`
3. Check `NAMESPACE`
4. Repeat until `NAMESPACE` is correct

???

- roxygen tags like `@includes`, `@export`, `@importFrom` generate `NAMESPACE` und `Collate`

- for packages with many variables to export it may be more convenient to specify the names to export with a regular expression
```{r}
#exportPattern("\^{}[\^{}\textbackslash\textbackslash.]")
```

---
class: inverse
# Workshop Outline: Part II

Basic R package components: 

- *DESCRIPTION* - What does the package? 

- *R/* - Start to write R functions, 

- *data/* - integrate examples and data, 

- *man/* - document them with help pages, 

- #### *tests/* - make sure that your code does what is expected by implementing unit testing, and 

- finally built it and put it on CRAN or wherever you want to share it. 

---
# Testing

"A unit testing system designed to be fun, exible and easy to set up." (Wickham)

- Provides functions that make it easy to describe what you expect a function to do, including catching errors, warnings
and messages.

- Displays test progress visually, showing a pass, fail or error for
every expectation. If you're using the terminal, it'll even
colour the output.

```{r echo=TRUE, eval=FALSE}
library(testthat)
library(yourpackage)

test_check("yourpackage")
```

---
# Testing

- `expect_that` describes expected result of your code (value, class, correct error message, computation time, etc.)
- `test_that` is grouping a number of expectation's for one functions or a feature
- `context()` is grouping a number of content-related tests

```{r echo=TRUE, eval=FALSE}
require(testthat)
test_that("trigonometric functions match identities", {
expect_that(sin(pi / 4), equals(1 / sqrt(2)))
expect_that(cos(pi / 4), equals(1 / sqrt(2)))
expect_that(tan(pi / 4), equals(1))
})
```

testthat: Hadley Wickham (2011). testthat: Get Started with Testing. The R Journal 3(1).

---
class: inverse
# Workshop Outline: Part II

Basic R package components: 

- *DESCRIPTION* - What does the package? 

- *R/* - Start to write R functions, 

- *data/* - integrate examples and data, 

- *man/* - document them with help pages, 

- *tests/* - make sure that your code does what is expected by implementing unit testing, and 

- #### finally built it and put it on CRAN or wherever you want to share it. 

---
# Building an R Package

For building a R package pkg run the following commands in your
console:

- `R CMD SHLIB pkg` compiles C/C++/Fortran code in pkg/src

- `R CMD build pkg` generates package bundle `pkg.tar.gz` or `pkg.zip`

- `R CMD INSTALL pkg.tar.gz` installs package

- `R CMD check pkg.tar.gz` runs CRAN validity checks (is pkg valid?)

**Note!** In windows, installation of `Rtools` is required.

---
#Example: Build and Check myutils

.small[
``` {r, eval=FALSE, echo=TRUE}
jmanitz@console$ R CMD build myutils_complete
* checking for file `myutils_complete/DESCRIPTION' ... OK
* preparing `myutils':
* checking DESCRIPTION meta-information ... OK
* installing the package to process help pages
* [...]
* building `myutils_1.0.tar.gz'

jmanitz@console$ R CMD check myutils_1.0.tar.gz
* using log directory `/home/Rladies/example/myutils.Rcheck'
* using R version 3.5.2 (2018-12-20)
* using platform: x86_64-pc-linux-gnu (64-bit)
* checking for file `myutils/DESCRIPTION' ... OK
* [...]
* checking PDF version of manual ... OK
* DONE
Status: OK
```
]

---
# R Packaging using `devtools`

- `load all()` simulates installing and reloading your package
--


- `document()` updates documentation, file collation and NAMESPACE.
--


- `check doc()` runs most of the documentation checking components of R CMD check
--


- `run_examples()` will run all examples to make sure they work.
--


- `test()` reloads your code, then runs all testthat tests.
--


- `check()` updates the documentation, then builds and checks the package
--


- `build()`, `build_win()` builds a package file from package sources (only one R version)

???
R functions from `devtools` that simplify R packaging:


---
# Other Package Elements

Standard structure

- `DESCRIPTION` what does the package? who can use it (license)? who is responsible (maintainer)?
- `NAMESPACE` which function should be seen by the user? which are internal?
- `R/` Write R functions
- `man/` documentation, help files with syntax similar to LATEX
- `data/` example data files
- `tests/` tests

Additional (optional) files in R packages:

- `vignettes/` vignette
- `doc/` pkgdown
- `inst/CITATION` how should the user cite the package?
- `src/` C, C++, FORTRAN source code

---
class: inverse
# Workshop Description: Part III

[...]

We conclude the workshop with a selection of advanced topics related to R packages (e.g., package validation, objected-oriented programming/S3 methods or advanced package documentation with `pkgdown`). 

--

1. Advanced package documentation 

2. Package validation 

3. Objected-oriented programming with S3 methods 

---
class: inverse
# Advanced Documentation [Leigh]

5 slides = 10mins

---
# Advanced Documentation

- vignettes
- websites

---
# Vignettes

- A vignette is a long-form guide to your package like a book chapter or an academic paper. 

- When writing a vignette, you’re teaching someone how to use your package.

- Data for vignettes: If you want to show how to work with an already loaded dataset, put that data in data/. If you want to show how to load raw data, put that data in inst/extdata. 

???

Function documentation is great if you know the name of the function you need, but it’s useless otherwise.

---
# pkgdown

- websites, 

---
class: inverse
# Package validation [Juliane]

(topic introduction: 3 slides + references)
5-10mins

---


---
class: inverse
# Object-Oriented Progamming (OOP) [Leigh/Juliane]

10slides 20min

---
# Object-Oriented Progamming 

- In Object-Oriented Progamming (OOP), computer programs are designed by making them out of objects that interact with one another

- a class defines the behaviour of objects by describing their attributes and their relationship to other classes.

- the class is also used when selecting methods, functions that behave differently depending on the class of their input.
- R has three OO systems: S3, S4, Reference classes, and the system of base types

---
# Picking an OOP System in R

- majority of object-oriented code that I have written in R is S3

- S3 is suffcient for fairly simple objects and methods for pre-existing generic functions like `print()`, `summary()`, and
`plot()`
- S4/RC may be more appropriate for more complicated systems of interrelated objects

- follow good examples from other packages: 

  - for S4 is the `Matrix` package by Douglas Bates and Martin Maechler

  - for RC is the `dplyr` package by RStudio

---
# Object-Oriented Progamming (OOP) using S3

- roxygen tags like `@method`, `@importClassesFrom`, `@importMethodsFrom` are used for for OOP documentation


